Variance Adaptive Shrinkage (vash): Flexible Empirical Bayes estimation of variances

### 2016-04-13 paper/main.tex
edit paper
tweak figures, add makefiles

\subsection{A unimodal distribution for the variances}
Let $\IG(\cdot;a,b)$ denote the density of an Inverse-Gamma distribution with shape $a$ and rate $b$. This distribution is unimodal with mode
at $c=b/(a+1)$. To obtain a more flexible family of unimodal distributions with mode at $c$ we consider a mixture of
Inverse-Gamma distributions, each with mode at $c$:
\begin{equation}
g(\cdot ; \piv,\av,  c) = \sum\limits_{k=1}^K \pi_{k} \IG(\cdot;a_k,b_k), \label{eqn:h}
\end{equation}
where
\begin{equation}
b_k:=(a_k+1)*c. \label{eqn:b1}
\end{equation}
Each component in (\ref{eqn:h}) has mode at $c$, and the variance about this mode is controlled by $a_k$, with large $a_k$
corresponding to small variance. By setting $\av$ to a fixed grid of values that range from ``small" to ``large", we obtain a flexible
family of distributions, with hyperparameters $\piv$, that are unimodal about $c$. See below for details on choice of grid for $\av$.
This approach is analogous to \citet{stephens2016ash}, which uses mixtures of normal or uniform distributions, with a fixed grid of variances,
to model unimodal distributions for mean parameters.

@@

For $K>1$, as noted above, we use $K$ ``large" (e.g. 10-16), fix the values of $a_k$ to a grid of values from ``small" to ``large", and estimate the hyper-parameters $c,\piv$ by maximizing the likelihood
\begin{align}
L(\av, c; \hat{s}_1^2,..., \hat{s}_G^2) &=P(\hat{s}_1,...,\hat{s}_G| \av, c) \\ 
&=\prod\limits_{g=1}^G \sum_k \pi_k p(\hat{s}_g; a_k,c) \label{eqn:lik}
\end{align}
where $p(\hat{s}_g; a_k,c)$ is given by (\ref{eqn:ps}).
 We center the grid of $a_k$ values on the point estimate $\hat{a}$ obtained for $K=1$, to ensure that the grid values span a range consistent with the data. Moreover, if the data are consistent with $K=1$ then the estimated $\piv$ will be concentrated on the component with $a_k=\hat{a}$, and thus lead to similar results to limma.

@@

\subsection{Posterior calculations}
Using (\ref{eqn:h}) as a prior distribution for $s_g^2$, and combining with the likelihood (\ref{eqn:selik}) the posterior distribution of $s_g^2$ is also a mixture of inverse Gamma distributions:
\begin{align}
P(s_g^2|\hat{s}_g^2) %&= \frac{P(\hat{s}_g^2|s_g^2)P(s_g^2|c,\av,\piv)}{\int P(\hat{s}_g^2|s_g^2)P(s_g^2|c,\av,\piv)ds_g}\\
& =\sum_k \tilde{\pi}_{gk} \IG(s_g^2; \tilde{a}_{gk}, \tilde{b}_{gk}),
\end{align}
where
\begin{align}
\tilde{a}_{gk}&:=a_k+d_g/2 , \label{post}  \\
\tilde{b}_{gk}&:=b_k+d_g \hat{s}_g^2/2, \\
\tilde{\pi}_{gk}& := \frac{\pi_k \hat{s}_g^{d_g-2}\frac{\Gamma(a_k+d_g/2)}{\Gamma(a_k)}\frac{b_k^{a_k}}{(b_k+d_g\hat{s}_g^2/2)^{a_k+d_g/2}}}
{\sum_{k'} \pi_{k'} \hat{s}_g^{d_g-2}\frac{\Gamma(a_{k'}+d_g/2)}{\Gamma(a_{k'})}\frac{b_{k'}^{a_{k'}}}{(b_{k'}+d_g\hat{s}_g^2/2)^{a_{k'}+d_g/2}}}.   \label{eqn:postpi}
\end{align}

@@

The above formulation is based on assuming a unimodal prior distribution for the variance $s_g^2$,
and specifically by using a mixture of inverse gamma distributions all with the same mode.
An alternative is to assume a unimodal prior distribution for the precision $1/s_g^2$,
by using a mixture of Gamma distributions, all with the same mode. 
This is equivalent to using a mixture
of inverse gamma distributions for the variance $s_g^2$ as in (\ref{eqn:h}) above, but with
\begin{equation}
b_k:= (a_k-1)/c \label{eqn:b2}
\end{equation}
in place of (\ref{eqn:b1}), because the mode of a $\Gam(a,b)$ distribution is at $c=(a-1)/b$.
We present results for both approaches. In practice one can assess which of the two models provides
a better fit to the data by comparing their (maximized) likelihoods (\ref{eqn:lik}).
 Note that in many (but not all) cases the fitted prior distributions under either or both approaches 
will end up being unimodal for both the variance {\it and} the precision. However, even in these cases,
the optimal likelihood under each approach will typically differ because the family of unimodal distributions 
being optimized over is different.

@@

To compare and contrast our method with limma and limmaR we simulate data from the model (\ref{eqn:selik})-(\ref{eqn:h}), with $G=10,000$, and degrees of freedom $df=3,10,50$ (corresponding to sample sizes 4, 11 and 51 respectively)
under various scenarios for the actual distribution of variances (scenarios A-D) or precisions (scenarios E-H), as summarized in Tables \ref{tab:simparams1} and \ref{tab:simparams2}. 
%We assume that 90\% of the genes are not differential expressed ($\beta_g=0$), while the rest of the genes are ($\beta_g\sim N(0,\sigma^2)$). Here $\sigma$ is held fixed at 2. 
The simulation scenarios are designed to span the range from a single inverse-gamma prior as assumed by limma, to more complex distributions under
which we might expect our method to outperform limma. Specifically we consider:
\begin{itemize}
\item Single IG (Gamma): single component inverse-Gamma prior on variance (or Gamma prior on precision), which satisfies the assumptions of limma.
\item Single IG (Gamma) with outliers: two component inverse-gamma prior on variance (or Gamma prior on precision), where one component models the majority of genes and the other component, being more spread out, attempts to capture possible outliers. The method limmaR is specifically designed to deal with the case where large variance outliers exist.
\item IG (Gamma) mixture: a more flexible inverse-Gamma mixture prior on variance (or mixture Gamma prior on precision) with multiple components.
\item Long tail log-normal mixture: log-normal mixture prior on variance or precision, which yields a longer tail than either the inverse-Gamma or the Gamma distribution.
\end{itemize}

@@

\begin{table}[!hbp]
\begin{center}
\caption{Parameters for the simulation scenarios  with unimodal prior on precision}
\label{tab:simparams2}
\begin{tabular}{p{0.1\linewidth}p{0.35\linewidth}p{0.5\linewidth}} \hline
Scenario & Description & Prior of $1/s_g^2$ \\ \hline
E & Single Gamma  & \Gam(10,9)\\
F & Single Gamma with outliers & 0.1\Gam(2,1)+0.9\Gam(10,9)\\
G & Gamma mixture & 0.1\Gam(2,1) + 0.4\Gam(5,4) + 0.5\Gam(30,29)\\
H &Long tail log-normal mixture & 0.7\logN(0.0625,0.0625) + 0.3\logN(0.64,0.64)\\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{figure}[!hbp]
\includegraphics[width=\linewidth]{relmse2_prec.eps}
\caption{$RRMSE_{prec}$ of three gene-specific variances estimators, limma, robust limma (limmaR) and our proposed estimator (vash) in the 4 simulation scenarios E-H with unimodal precision prior.}
\label{fg:rrmse2}
\end{figure}

%In summary, vash performs very similarly to limma and limmaR when the true $g$ is a single inverse gamma distribution, while its additional modeling flexibility can yield improved accuracy in other scenarios.

@@

The results above demonstrate that the more flexible mixture prior implemented in vash, can in principle provide more accurate variance and precision estimates than
the simple inverse gamma prior implemented in limma. However, in practice these gains will only be realized if the actual distribution of variances differs from the single inverse gamma model. Here
we examine this issue using RNA sequencing data from the Genotype-Tissue Expression (GTEx) project \citep{gtex}. The GTEx Project is an extensive resource which studies the relationship among genetic variation, gene expression, and other molecular phenotypes in multiple human tissues. Here we consider RNA-seq data (GTEx V6 dbGaP accession phs000424.v6.p1, release date: Oct 19, 2015, \href{http://www.gtexportal.org/home/}{http://www.gtexportal.org/home/}) on 53 human tissues from a total of 8555 samples (ranging from 6 to 430 samples per tissues).
 %This sample size is sufficiently large to capture the variance (or precision) distribution with high accuracy.

@@

Our method is implemented in an R package \texttt{vashr} available from \href{http://github.com/mengyin/vashr}{http://github.com/mengyin/vashr}.

>>>>

\subsection{A unimodal distribution for the variances}
Let $\IG(\cdot;a,b)$ denote the density of an inverse-gamma distribution with shape $a$ and rate $b$. This distribution is unimodal with mode
at $c=b/(a+1)$. To obtain a more flexible family of unimodal distributions with mode at $c$ we consider a mixture of
inverse-gamma distributions, each with mode at $c$:
\begin{equation}
g(\cdot ; \piv,\av,  c) = \sum\limits_{k=1}^K \pi_{k} \IG(\cdot;a_k,b_k), \label{eqn:h}
\end{equation}
where
\begin{equation}
b_k:=(a_k+1)c. \label{eqn:b1}
\end{equation}
Each component in (\ref{eqn:h}) has mode at $c$, and the variance about this mode is controlled by $a_k$, with large $a_k$
corresponding to small variance. By setting $\av$ to a fixed grid of values that range from ``small" to ``large", we obtain a flexible
family of distributions, with hyperparameters $\piv$, that are unimodal about $c$. See below for details on choice of grid for $\av$.
This approach is analogous to \citet{stephens2016ash}, which uses mixtures of normal or uniform distributions, with a fixed grid of variances,
to model unimodal distributions for mean parameters.

@@

For $K>1$, as noted above, we use $K$ ``large" (e.g. 10-16), fix the values of $a_k$ to a grid of values from ``small" to ``large", and estimate the hyper-parameters $c,\piv$ by maximizing the likelihood
\begin{align}
L(\piv, c;\av, \hat{s}_1^2,..., \hat{s}_G^2) &=P(\hat{s}_1,...,\hat{s}_G| \piv,\av, c) \\ 
&=\prod\limits_{g=1}^G \sum_k \pi_k p(\hat{s}_g; a_k,c) \label{eqn:lik}
\end{align}
where $p(\hat{s}_g; a_k,c)$ is given by (\ref{eqn:ps}).
 We center the grid of $a_k$ values on the point estimate $\hat{a}$ obtained for $K=1$, to ensure that the grid values span a range consistent with the data. Moreover, if the data are consistent with $K=1$ then the estimated $\piv$ will be concentrated on the component with $a_k=\hat{a}$, and thus lead to similar results to limma.

@@

\subsection{Posterior calculations}
Using (\ref{eqn:h}) as a prior distribution for $s_g^2$, and combining with the likelihood (\ref{eqn:selik}) the posterior distribution of $s_g^2$ is also a mixture of inverse-gamma distributions:
\begin{align}
P(s_g^2|\hat{s}_g^2) %&= \frac{P(\hat{s}_g^2|s_g^2)P(s_g^2|c,\av,\piv)}{\int P(\hat{s}_g^2|s_g^2)P(s_g^2|c,\av,\piv)ds_g}\\
& =\sum_k \tilde{\pi}_{gk} \IG(s_g^2; \tilde{a}_{gk}, \tilde{b}_{gk}),
\end{align}
where
\begin{align}
\tilde{a}_{gk}&:=a_k+d_g/2 , \label{post}  \\
\tilde{b}_{gk}&:=b_k+d_g \hat{s}_g^2/2, \\
\tilde{\pi}_{gk}& := \frac{\pi_k \hat{s}_g^{d_g-2}\frac{\Gamma(a_k+d_g/2)}{\Gamma(a_k)}\frac{b_k^{a_k}}{(b_k+d_g\hat{s}_g^2/2)^{a_k+d_g/2}}}
{\sum_{k'} \pi_{k'} \hat{s}_g^{d_g-2}\frac{\Gamma(a_{k'}+d_g/2)}{\Gamma(a_{k'})}\frac{b_{k'}^{a_{k'}}}{(b_{k'}+d_g\hat{s}_g^2/2)^{a_{k'}+d_g/2}}}.   \label{eqn:postpi}
\end{align}

@@

The above formulation is based on assuming a unimodal prior distribution for the variance $s_g^2$,
and specifically by using a mixture of inverse-gamma distributions all with the same mode.
An alternative is to assume a unimodal prior distribution for the precision $1/s_g^2$,
by using a mixture of gamma distributions, all with the same mode. 
This is equivalent to using a mixture
of inverse-gamma distributions for the variance $s_g^2$ as in (\ref{eqn:h}) above, but with
\begin{equation}
b_k:= (a_k-1)/c \label{eqn:b2}
\end{equation}
in place of (\ref{eqn:b1}), because the mode of a $\Gam(a,b)$ distribution is at $c=(a-1)/b$.
We present results for both approaches. In practice one can assess which of the two models provides
a better fit to the data by comparing their (maximized) likelihoods (\ref{eqn:lik}).
 Note that in many (but not all) cases the fitted prior distributions under either or both approaches 
will end up being unimodal for both the variance {\it and} the precision. However, even in these cases,
the optimal likelihood under each approach will typically differ because the family of unimodal distributions 
being optimized over is different.

@@

To compare and contrast our method with limma and limmaR we simulate data from the model (\ref{eqn:selik})-(\ref{eqn:h}), with $G=10,000$, and degrees of freedom $df=3,10,50$ (corresponding to sample sizes 4, 11 and 51 respectively)
under various scenarios for the actual distribution of variances (scenarios A-D) or precisions (scenarios E-H), as summarized in Tables \ref{tab:simparams1} and \ref{tab:simparams2}. 
%We assume that 90\% of the genes are not differential expressed ($\beta_g=0$), while the rest of the genes are ($\beta_g\sim N(0,\sigma^2)$). Here $\sigma$ is held fixed at 2. 
The simulation scenarios are designed to span the range from a single inverse-gamma prior as assumed by limma, to more complex distributions under
which we might expect our method to outperform limma. Specifically we consider:
\begin{itemize}
\item Single IG (Gamma): single component inverse-gamma prior on variance (or gamma prior on precision), which satisfies the assumptions of limma.
\item Single IG (Gamma) with outliers: two component inverse-gamma prior on variance (or gamma prior on precision), where one component models the majority of genes and the other component, being more spread out, attempts to capture possible outliers. The method limmaR is specifically designed to deal with the case where large variance outliers exist.
\item IG (Gamma) mixture: a more flexible inverse-gamma mixture prior on variance (or mixture gamma prior on precision) with multiple components.
\item Long tail log-normal mixture: log-normal mixture prior on variance or precision, which yields a longer tail than either the inverse-gamma or the gamma distribution.
\end{itemize}

@@

\begin{table}[!hbp]
\begin{center}
\caption{Parameters for the simulation scenarios  with unimodal prior on precision}
\label{tab:simparams2}
\begin{tabular}{p{0.1\linewidth}p{0.35\linewidth}p{0.5\linewidth}} \hline
Scenario & Description & Prior of $1/s_g^2$ \\ \hline
E & Single gamma  & \Gam(10,9)\\
F & Single gamma with outliers & 0.1\Gam(2,1)+0.9\Gam(10,9)\\
G & Gamma mixture & 0.1\Gam(2,1) + 0.4\Gam(5,4) + 0.5\Gam(30,29)\\
H &Long tail log-normal mixture & 0.7\logN(0.0625,0.0625) + 0.3\logN(0.64,0.64)\\ \hline
\end{tabular}
\end{center}
\end{table}

@@

\begin{figure}[!hbp]
\includegraphics[width=\linewidth]{relmse2_prec.eps}
\caption{$\RRMSE_{prec}$ of three gene-specific variances estimators, limma, robust limma (limmaR) and our proposed estimator (vash) in the 4 simulation scenarios E-H with unimodal precision prior.}
\label{fg:rrmse2}
\end{figure}

%In summary, vash performs very similarly to limma and limmaR when the true $g$ is a single inverse-gamma distribution, while its additional modeling flexibility can yield improved accuracy in other scenarios.

@@

The results above demonstrate that the more flexible mixture prior implemented in vash, can in principle provide more accurate variance and precision estimates than
the simple inverse-gamma prior implemented in limma. However, in practice these gains will only be realized if the actual distribution of variances differs from the single inverse-gamma model. Here
we examine this issue using RNA sequencing data from the Genotype-Tissue Expression (GTEx) project \citep{gtex}. The GTEx Project is an extensive resource which studies the relationship among genetic variation, gene expression, and other molecular phenotypes in multiple human tissues. Here we consider RNA-seq data (GTEx V6 dbGaP accession phs000424.v6.p1, release date: Oct 19, 2015, \href{http://www.gtexportal.org/home/}{http://www.gtexportal.org/home/}) on 53 human tissues from a total of 8555 samples (ranging from 6 to 430 samples per tissues).
 %This sample size is sufficiently large to capture the variance (or precision) distribution with high accuracy.

@@

Our method is implemented in an R package \texttt{vashr} available from \href{http://github.com/mengyin/vashr}{http://github.com/mengyin/vashr}. Codes for reproducing analyses and figures in this paper are at \href{https://github.com/mengyin/vash}{https://github.com/mengyin/vash}.
